Geometric models reveal behavioral and neural signatures of transforming experiences into memories

### 2020-10-14 paper/main.tex
remove merge markers and recompile
reformatted to satisfy NHB requirements and did some minor wording edits throughout

\newcommand{\topicopt}{S1}
\newcommand{\topics}{S2}
\newcommand{\featureimportance}{S3}
\newcommand{\corrmats}{S4}
\newcommand{\matchmats}{S5}
\newcommand{\kopt}{S6}

@@

\section*{Introduction}
What does it mean to remember something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a discrete, binary operation: each studied item may be separated from the rest of one's experience and labeled as having been either recalled or forgotten. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between recollecting the (contextual) details of an experience and having a general feeling of familiarity~\citep{Yone02}. Using well-controlled, trial-based experimental designs, the field has amassed a wealth of information regarding human episodic memory~\citep[for review see][]{Kaha12}.  However, there are fundamental properties of the external world and our memories that trial-based experiments are not well suited to capture~\citep[for review, also see][]{KoriGold94, HukEtal18}.  First, our experiences and memories are continuous, rather than discrete---isolating a naturalistic event from the context in which it occurs can substantially change its meaning.  Second, whether or not the rememberer has precisely reproduced a specific set of words in describing a given experience is nearly orthogonal to how well they were actually able to remember it.  In classic (e.g., list-learning) memory studies, by contrast, the number or proportion of exact recalls is often considered to be a primary metric for assessing the quality of participants' memories.  Third, one might remember the essence (or a general summary) of an experience but forget (or neglect to recount) particular low-level details.  Capturing the essence of what happened is often a main goal of recounting an episodic memory to a listener, whereas the inclusion of specific low-level details is often less pertinent.

How might we formally characterize the ``essence'' of an experience, and whether it has been recovered by the rememberer?  And how might we distinguish an experience's overarching essence from its low-level details?  One approach is to start by considering some fundamental properties of the dynamics of our experiences.  Each given moment of an experience tends to derive meaning from surrounding moments, as well as from longer-range temporal associations~\citep{LernEtal11, Mann19, Mann20}.  Therefore, the timecourse describing how an event unfolds is fundamental to its overall meaning.  Further, this hierarchy formed by our subjective experiences at different timescales defines a context for each new moment~\citep[e.g.,][]{HowaKaha02a, HowaEtal14}, and plays an important role in how we interpret that moment and remember it later~\citep[for review see][]{MannEtal15, Mann20}.  Our memory systems can leverage these associations to form predictions that help guide our behaviors~\citep{RangRitc12}.  For example, as we navigate the world, the features of our subjective experiences tend to change gradually (e.g., the room or situation we find ourselves in at any given moment is strongly temporally autocorrelated), allowing us to form stable estimates of our current situation and behave accordingly~\citep{ZackEtal07, ZwaaRadv98}.

Occasionally, this gradual drift of our ongoing experience is punctuated by sudden changes, or shifts~\citep[e.g., when we walk through a doorway;][]{RadvZack17}.  Prior research suggests that these sharp transitions (termed ``event boundaries'') help to discretize our experiences (and their mental representations) into ``events''~\citep{RadvZack17, BrunEtal18, HeusEtal18b, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within-event) and transient (across-event) temporal dynamics of an experience also provides a potential framework for transforming experiences into memories that distills those experiences down to their essences.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18b, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}.  This work also suggests a means of distinguishing the essence of an experience from its low-level details:  The overall structure of events and event transitions reflects how the high-level experience unfolds (i.e., its essence), while subtler event-level properties reflect its low-level details.  Prior research has also implicated a network of brain regions (including the hippocampus and the medial prefrontal cortex) in playing a critical role in transforming experiences into structured and consolidated memories ~\citep{TompDava17}.

@@

\cite{ChenEtal17} collected hand-annotated information about each of 1,000 (manually delineated) time segments spanning the roughly 50 minute video used in their study.  Each annotation included: a brief narrative description of what was happening, the location where the action took place, the names of any characters on the screen, and other similar details (for a full list of annotated features, see \textit{Methods}).  We took the union of all unique words (excluding stop words, such as ``and,'' ``or,'' ``but,'' etc.) across all features from all annotations as the vocabulary for the topic model.  We then concatenated the sets of words across all features contained in overlapping sliding windows of (up to) 50 annotations, and treated each window as a single document for the purpose of fitting the topic model.  Next, we fit a topic model with (up to) $K = 100$ topics to this collection of documents.  We found that 32 unique topics (with non-zero weights) were sufficient to describe the time-varying content of the episode (see \textit{Methods}; Fig.~\ref{fig:schematic}, Supp. Fig.~\topics).  We note that our approach is similar in some respects to Dynamic Topic Models~\citep{BleiLaff06} in that we sought to characterize how the thematic content of the episode evolved over time.  However, whereas Dynamic Topic Models are designed to characterize how the properties of collections of documents change over time, our sliding window approach allows us to examine the topic dynamics within a single document (or video).  Specifically, our approach yielded (via the topic proportions matrix) a single ``topic vector'' for each sliding window of annotations transformed by the topic model.  We then stretched (interpolated) the resulting windows-by-topics matrix to match the time series of the 1,976 fMRI volumes collected as participants viewed the episode.

@@

The 32 topics we found were heavily character-focused (i.e., the top-weighted word in each topic was nearly always a character) and could be roughly divided into themes centered around Sherlock Holmes (the titular character), John Watson (Sherlock's close confidant and assistant), supporting characters (e.g., Inspector Lestrade, Sergeant Donovan, or Sherlock's brother Mycroft), or the interactions between various groupings of these characters (Supp. Fig.~\topics).  This likely follows from the frequency with which these terms appeared in the episode annotations.  Several of the identified topics were highly similar, which we hypothesized might allow us to distinguish between subtle narrative differences if the distinctions between those overlapping topics were meaningful.  The topic vectors for each timepoint were also sparse, in that only a small number of topics (typically one or two) tended to be ``active'' in any given timepoint (Fig.~\ref{fig:model}A).  Further, the dynamics of the topic activations appeared to exhibit persistence (i.e., given that a topic was active in one timepoint, it was likely to be active in the following timepoint) along with occasional rapid changes (i.e., occasionally topic weights would change abruptly from one timepoint to the next).  These two properties of the topic dynamics may be seen in the block diagonal structure of the timepoint-by-timepoint correlation matrix (Fig.~\ref{fig:model}B) and reflect the gradual drift and sudden shifts fundamental to the temporal dynamics of many real-world experiences, as well as television episodes.  Given this observation, we adapted an approach devised by \cite{BaldEtal17}, and used a hidden Markov model (HMM) to identify the ``event boundaries'' where the topic activations changed rapidly (i.e., the boundaries of the blocks in the temporal correlation matrix; event boundaries identified by the HMM are outlined in yellow in Fig.~\ref{fig:model}B).  Part of our model fitting procedure required selecting an appropriate number of events into which the topic trajectory should be segmented.  To accomplish this, we used an optimization procedure that maximized the difference between the topic weights for timepoints within an event versus timepoints across multiple events (see \textit{Methods}).  We then created a stable summary of the content within each episode event by averaging the topic vectors across the timepoints spanned by each event (Fig.~\ref{fig:model}C).

@@

Although the example participant's recall topic proportions matrix has some visual similarity to the episode topic proportions matrix, the time-varying topic proportions for the example participant's recalls are not as sparse as those for the episode (compare Figs.~\ref{fig:model}A and D).  Similarly, although there do appear to be periods of stability in the recall topic dynamics (i.e., most topics are active or inactive over contiguous blocks of time), the changes in topic activations that define event boundaries appear less clearly delineated in participants' recalls than in the episode's annotations.  To examine these patterns in detail, we computed the timepoint-by-timepoint correlation matrix for the example participant's recall topic proportions matrix (Fig.~\ref{fig:model}E).  As in the episode correlation matrix (Fig.~\ref{fig:model}B), the example participant's recall correlation matrix has a strong block diagonal structure, indicating that their recalls are discretized into separated events.  We used the same HMM-based optimization procedure that we had applied to the episode's topic proportions matrix (see \textit{Methods}) to estimate an analogous set of event boundaries in the participant's recounting of the episode (outlined in yellow).  We carried out this analysis on all 17 participants' recall topic proportions matrices (Supp. Fig.~\corrmats).

Two clear patterns emerged from this set of analyses.  First, although every individual participant's recalls could be segmented into discrete events (i.e., every individual participant's recall correlation matrix exhibited clear block diagonal structure; Supp. Fig.~\corrmats), each participant appeared to have a unique ``recall resolution,'' reflected in the sizes of those blocks.  While some participants' recall topic proportions segmented into just a few events (e.g., Participants P4, P5, and P7), others' segmented into many shorter-duration events (e.g., Participants P12, P13, and P17).  This suggests that different participants may be recalling the episode with different levels of detail---i.e., some might recount only high-level essential plot details, whereas others might recount low-level details instead (or in addition).  The second clear pattern present in every individual participant's recall correlation matrix was that, unlike in the episode correlation matrix, there were substantial off-diagonal correlations.  One potential explanation for this finding is that the topic models, trained only on episode annotations, do not capture topic proportions in participants' ``held-out'' recalls as accurately.  A second possibility is that, whereas each event in the original episode was (largely) separable from the others (Fig.~\ref{fig:model}B), in transforming those separable events into memory, participants appeared to be integrating across multiple events, blending elements of previously recalled and not-yet-recalled content into each newly recalled event~\citep[Fig.~\ref{fig:model}E, Supp. Fig.~\corrmats; also see][]{MannEtal11, HowaEtal12, Mann19}.

The above results demonstrate that topic models capture the dynamic conceptual content of the episode and participants' recalls of the episode.  Further, the episode and recalls exhibit event boundaries that can be identified automatically using HMMs to segment the dynamic content.  Next, we asked whether some correspondence might be made between the specific content of the events the participants experienced while viewing the episode, and the events they later recalled.  We labeled each recall event as matching the episode event with the most similar (i.e., most highly correlated) topic vector (Fig.~\ref{fig:model}G, Supp. Fig.~\matchmats).  This yielded a sequence of ``presented'' events from the original episode, and a (potentially differently ordered) sequence of ``recalled'' events for each participant.  Analogous to classic list-learning studies, we can then examine participants' recall sequences by asking which events they tended to recall first~\citep[probability of first recall; Fig.~\ref{fig:list-learning}A;][]{AtkiShif68, PostPhil65, WelcBurn24}; how participants most often transitioned between recalls of the events as a function of the temporal distance between them~\citep[lag-conditional response probability; Fig.~\ref{fig:list-learning}B;][]{Kaha96}; and which events they were likely to remember overall~\citep[serial position recall analyses; Fig.~\ref{fig:list-learning}C;][]{Murd62a}. Some of the patterns we observed appeared to be similar to classic effects from the list-learning literature.  For example, participants had a higher probability of initiating recall with early events (Fig.~\ref{fig:list-learning}A) and a higher probability of transitioning to neighboring events with an asymmetric forward bias (Fig.~\ref{fig:list-learning}B). However, unlike what is typically observed in list-learning studies, we did not observe patterns comparable to the primacy or recency serial position effects (Fig.~\ref{fig:list-learning}C).  We hypothesized that participants might be leveraging meaningful narrative associations and references over long timescales throughout the episode.

Clustering scores are often used by memory researchers to characterize how people organize their memories of words on a studied list~\citep[for review, see][]{PolyEtal09}.  We defined analogous measures to characterize how participants organized their memories for episodic events (see \textit{Methods} for details).  Temporal clustering refers to the extent to which participants group their recall responses according to encoding position.  Overall, we found that sequentially viewed episode events tended to appear nearby in participants' recall event sequences (mean clustering score: 0.732, SEM: 0.033).  Participants with higher temporal clustering scores tended to exhibit better overall memory for the episode, according to both \cite{ChenEtal17}'s hand-counted numbers of recalled scenes from the episode (Pearson's $r(15) = 0.49,~p = 0.046$) and the numbers of episode events that best-matched at least one recall event (i.e., model-estimated number of events recalled; Pearson's $r(15) = 0.59,~p = 0.013$).  Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity.  We found that participants tended to recall semantically similar episode events together (mean clustering score: 0.650, SEM: 0.032), and that semantic clustering score was also related to both hand-counted  (Pearson's $r(15) = 0.65,~p = 0.005$) and model-estimated (Pearson's $r(15) = 0.58,~p = 0.015$) numbers of recalled events.

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_distinctiveness}
  \caption{\small \textbf{Novel content-based metrics of naturalistic memory: precision and distinctiveness.} \textbf{A.} The episode-recall correlation matrix for a representative participant (P17), chosen for their large number of recall events (for analogous figures for other participants, see Supp. Fig.~\corrmats).  The yellow boxes highlight the maximum correlation in each column.  The example participant's overall precision score was computed as the average across the (Fisher $z$-transformed) correlation values in the yellow boxes.  Their distinctiveness score was computed as the average (over recall events) of the $z$-scored (within column) event precisions. \textbf{B.} The (Pearson's) correlation between precision and hand-counted number of recalled scenes.  \textbf{C.} The correlation between distinctiveness and hand-counted number of recalled scenes. \textbf{D.} The correlation between precision and the number of recalled episode events, as determined by our model. \textbf{E.} The correlation between distinctiveness and the number of recalled episode events, as determined by our model.}
  \label{fig:precision-distinctiveness}
\end{figure}

@@

\begin{figure}[tp]
  \centering
  \vspace*{-1cm}
  \includegraphics[width=.7\textwidth]{figs/precision_distinctiveness_detail}
  \caption{\small \textbf{Precision reflects the completeness of recall, whereas distinctiveness reflects recall specificity.} \textbf{A.} Recall precision by episode event.  Grey violin plots display kernel density estimates for the distribution of recall precision scores for a single episode event.  Colored dots within each violin plot represent individual participants' recall precision for the given event.  \textbf{B.} Recall distinctiveness by episode event, analogous to Panel A.  \textbf{C.} The set of ``Narrative Details'' episode annotations (generated by \citealp{ChenEtal17}) comprising an example episode event (22) identified by the HMM.  Each action or feature is highlighted in a different color.  \textbf{D.} Sentences comprising the most precise (P17) and least precise (P6) participants' recalls of episode event 21.  Descriptions of specific actions or features reflecting those highlighted in Panel B are highlighted in the corresponding color.  The text highlighted in gray denotes a (rare) false recall.  The unhighlighted text denotes correctly recalled information about other episode events.  \textbf{E.} The sets of ``Narrative Details'' episode annotations (generated by \citealp{ChenEtal17}) for scenes comprising episode events described by the example participants in Panel F.  Each event's text is highlighted in a different color.  \textbf{F.} The sentences comprising the most distinctive (P9) and least distinctive (P6) participants' recalls of episode event 21.  Sections of recall describing each each episode event in Panel E are highlighted with the corresponding color.}
  \label{fig:precision-detail}
\end{figure}

@@

The preceding analyses sought to characterize how participants' recountings of individual episode events captured the low-level details of each event.  Next, we sought to characterize how participants' recountings of the full episode captured its high-level essence---i.e., the shape of the episode's trajectory through word embedding (topic) space.  To visualize the essence of the episode and each participant's recall trajectory~\citep{HeusEtal18a}, we projected the topic proportions matrices for the episode and recalls onto a shared two-dimensional space using Uniform Manifold Approximation and Projection~\citep[UMAP; ][]{McInEtal18}.  In this lower-dimensional space, each point represents a single episode or recall event, and the distances between the points reflect the distances between the events' associated topic vectors (Fig.~\ref{fig:trajectory}). In other words, events that are nearer to each other in this space are more semantically similar, and those that are farther apart are less so.

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/trajectory}
\caption{\small \textbf{Trajectories through topic space capture the dynamic content of the episode and recalls.}  All panels: the topic proportion matrices have been projected onto a shared two-dimensional space using UMAP.  \textbf{A.} The two-dimensional topic trajectory taken by the episode of \textit{Sherlock}.  Each dot indicates an event identified using the HMM (see \textit{Methods}); the dot colors denote the order of the events (early events are in red; later events are in blue), and the connecting lines indicate the transitions between successive events.  \textbf{B.} The average two-dimensional trajectory captured by participants' recall sequences, with the same format and coloring as the trajectory in Panel A.  To compute the event positions, we matched each recalled event with an event from the original episode (see \textit{Results}), and then we averaged the positions of all events with the same label.  The arrows reflect the average transition direction through topic space taken by any participants whose trajectories crossed that part of topic space; blue denotes reliable agreement across participants via a Rayleigh test ($p < 0.05$, corrected).  \textbf{C.} The recall topic trajectories (gray) taken by each individual participant (P1--P17).  The episode's trajectory is shown in black for reference.  Here, events (dots) are colored by their matched episode event (Panel A).}
\label{fig:trajectory}
\end{figure}

>>>>

\newcommand{\topicopt}{1}
\newcommand{\topics}{2}
\newcommand{\featureimportance}{3}
\newcommand{\corrmats}{4}
\newcommand{\matchmats}{5}
\newcommand{\kopt}{6}
\newcommand{\arrows}{7}

@@

\section*{Introduction}
What does it mean to remember something? In traditional episodic memory experiments (e.g., list-learning or trial-based experiments\citep{Murd62a, Kaha96}), remembering is often cast as a discrete, binary operation: each studied item may be separated from the rest of one's experience and labeled as having been either recalled or forgotten. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between recollecting the (contextual) details of an experience and having a general feeling of familiarity~\citep{Yone02}. Using well-controlled, trial-based experimental designs, the field has amassed a wealth of information regarding human episodic memory~\citep{Kaha12}.  However, there are fundamental properties of the external world and our memories that trial-based experiments are not well suited to capture~\citep{KoriGold94, HukEtal18}.  First, our experiences and memories are continuous, rather than discrete---isolating a naturalistic event from the context in which it occurs can substantially change its meaning.  Second, whether or not the rememberer has precisely reproduced a specific set of words in describing a given experience is nearly orthogonal to how well they were actually able to remember it.  In classic (e.g., list-learning) memory studies, by contrast, the number or proportion of exact recalls is often considered to be a primary metric for assessing the quality of participants' memories.  Third, one might remember the essence (or a general summary) of an experience but forget (or neglect to recount) particular low-level details.  Capturing the essence of what happened is often a main goal of recounting an episodic memory to a listener, whereas the inclusion of specific low-level details is often less pertinent.

How might we formally characterize the ``essence'' of an experience, and whether it has been recovered by the rememberer?  And how might we distinguish an experience's overarching essence from its low-level details?  One approach is to start by considering some fundamental properties of the dynamics of our experiences.  Each given moment of an experience tends to derive meaning from surrounding moments, as well as from longer-range temporal associations~\citep{LernEtal11, Mann19, Mann20}.  Therefore, the timecourse describing how an event unfolds is fundamental to its overall meaning.  Further, this hierarchy formed by our subjective experiences at different timescales defines a context for each new moment~\citep{HowaKaha02a, HowaEtal14}, and plays an important role in how we interpret that moment and remember it later~\citep{MannEtal15, Mann20}.  Our memory systems can leverage these associations to form predictions that help guide our behaviors~\citep{RangRitc12}.  For example, as we navigate the world, the features of our subjective experiences tend to change gradually (e.g., the room or situation we find ourselves in at any given moment is strongly temporally autocorrelated), allowing us to form stable estimates of our current situation and behave accordingly~\citep{ZackEtal07, ZwaaRadv98}.

Occasionally, this gradual drift of our ongoing experience is punctuated by sudden changes, or shifts (e.g., when we walk through a doorway~\citep{RadvZack17}).  Prior research suggests that these sharp transitions (termed ``event boundaries'') help to discretize our experiences (and their mental representations) into ``events''~\citep{RadvZack17, BrunEtal18, HeusEtal18b, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within-event) and transient (across-event) temporal dynamics of an experience also provides a potential framework for transforming experiences into memories that distills those experiences down to their essences.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18b, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}.  This work also suggests a means of distinguishing the essence of an experience from its low-level details:  The overall structure of events and event transitions reflects how the high-level experience unfolds (i.e., its essence), while subtler event-level properties reflect its low-level details.  Prior research has also implicated a network of brain regions (including the hippocampus and the medial prefrontal cortex) in playing a critical role in transforming experiences into structured and consolidated memories ~\citep{TompDava17}.

@@

Chen et al. (2017) collected hand-annotated information about each of 1,000 (manually delineated) time segments spanning the roughly 50 minute video used in their study~\cite{ChenEtal17}.  Each annotation included: a brief narrative description of what was happening, the location where the action took place, the names of any characters on the screen, and other similar details (for a full list of annotated features, see \textit{Methods}).  We took the union of all unique words (excluding stop words, such as ``and,'' ``or,'' ``but,'' etc.) across all features from all annotations as the vocabulary for the topic model.  We then concatenated the sets of words across all features contained in overlapping sliding windows of (up to) 50 annotations, and treated each window as a single document for the purpose of fitting the topic model.  Next, we fit a topic model with (up to) $K = 100$ topics to this collection of documents.  We found that 32 unique topics (with non-zero weights) were sufficient to describe the time-varying content of the episode (see \textit{Methods}; Fig.~\ref{fig:schematic}, Supp.\ Fig.~\topics).  We note that our approach is similar in some respects to Dynamic Topic Models~\citep{BleiLaff06} in that we sought to characterize how the thematic content of the episode evolved over time.  However, whereas Dynamic Topic Models are designed to characterize how the properties of collections of documents change over time, our sliding window approach allows us to examine the topic dynamics within a single document (or video).  Specifically, our approach yielded (via the topic proportions matrix) a single ``topic vector'' for each sliding window of annotations transformed by the topic model.  We then stretched (interpolated) the resulting windows-by-topics matrix to match the time series of the 1,976 fMRI volumes collected as participants viewed the episode.

@@

The 32 topics we found were heavily character-focused (i.e., the top-weighted word in each topic was nearly always a character) and could be roughly divided into themes centered around Sherlock Holmes (the titular character), John Watson (Sherlock's close confidant and assistant), supporting characters (e.g., Inspector Lestrade, Sergeant Donovan, or Sherlock's brother Mycroft), or the interactions between various groupings of these characters (Supp.\ Fig.~\topics).  This likely follows from the frequency with which these terms appeared in the episode annotations.  Several of the identified topics were highly similar, which we hypothesized might allow us to distinguish between subtle narrative differences if the distinctions between those overlapping topics were meaningful.  The topic vectors for each timepoint were also sparse, in that only a small number of topics (typically one or two) tended to be ``active'' in any given timepoint (Fig.~\ref{fig:model}A).  Further, the dynamics of the topic activations appeared to exhibit persistence (i.e., given that a topic was active in one timepoint, it was likely to be active in the following timepoint) along with occasional rapid changes (i.e., occasionally topic weights would change abruptly from one timepoint to the next).  These two properties of the topic dynamics may be seen in the block diagonal structure of the timepoint-by-timepoint correlation matrix (Fig.~\ref{fig:model}B) and reflect the gradual drift and sudden shifts fundamental to the temporal dynamics of many real-world experiences, as well as television episodes.  Given this observation, we adapted an approach devised by Baldassano et al. (2017)~\cite{BaldEtal17}, and used a hidden Markov model (HMM) to identify the ``event boundaries'' where the topic activations changed rapidly (i.e., the boundaries of the blocks in the temporal correlation matrix; event boundaries identified by the HMM are outlined in yellow in Fig.~\ref{fig:model}B).  Part of our model fitting procedure required selecting an appropriate number of events into which the topic trajectory should be segmented.  To accomplish this, we used an optimization procedure that maximized the difference between the topic weights for timepoints within an event versus timepoints across multiple events (see \textit{Methods}).  We then created a stable summary of the content within each episode event by averaging the topic vectors across the timepoints spanned by each event (Fig.~\ref{fig:model}C).

@@

Although the example participant's recall topic proportions matrix has some visual similarity to the episode topic proportions matrix, the time-varying topic proportions for the example participant's recalls are not as sparse as those for the episode (compare Figs.~\ref{fig:model}A and D).  Similarly, although there do appear to be periods of stability in the recall topic dynamics (i.e., most topics are active or inactive over contiguous blocks of time), the changes in topic activations that define event boundaries appear less clearly delineated in participants' recalls than in the episode's annotations.  To examine these patterns in detail, we computed the timepoint-by-timepoint correlation matrix for the example participant's recall topic proportions matrix (Fig.~\ref{fig:model}E).  As in the episode correlation matrix (Fig.~\ref{fig:model}B), the example participant's recall correlation matrix has a strong block diagonal structure, indicating that their recalls are discretized into separated events.  We used the same HMM-based optimization procedure that we had applied to the episode's topic proportions matrix (see \textit{Methods}) to estimate an analogous set of event boundaries in the participant's recounting of the episode (outlined in yellow).  We carried out this analysis on all 17 participants' recall topic proportions matrices (Supp.\ Fig.~\corrmats).

Two clear patterns emerged from this set of analyses.  First, although every individual participant's recalls could be segmented into discrete events (i.e., every individual participant's recall correlation matrix exhibited clear block diagonal structure; Supp.\ Fig.~\corrmats), each participant appeared to have a unique ``recall resolution,'' reflected in the sizes of those blocks.  While some participants' recall topic proportions segmented into just a few events (e.g., Participants P4, P5, and P7), others' segmented into many shorter-duration events (e.g., Participants P12, P13, and P17).  This suggests that different participants may be recalling the episode with different levels of detail---i.e., some might recount only high-level essential plot details, whereas others might recount low-level details instead (or in addition).  The second clear pattern present in every individual participant's recall correlation matrix was that, unlike in the episode correlation matrix, there were substantial off-diagonal correlations.  One potential explanation for this finding is that the topic models, trained only on episode annotations, do not capture topic proportions in participants' ``held-out'' recalls as accurately.  A second possibility is that, whereas each event in the original episode was (largely) separable from the others (Fig.~\ref{fig:model}B), in transforming those separable events into memory, participants appeared to be integrating across multiple events, blending elements of previously recalled and not-yet-recalled content into each newly recalled event (Fig.~\ref{fig:model}E, Supp.\ Fig.~\corrmats)~\citep{MannEtal11, HowaEtal12, Mann19}.

The above results demonstrate that topic models capture the dynamic conceptual content of the episode and participants' recalls of the episode.  Further, the episode and recalls exhibit event boundaries that can be identified automatically using HMMs to segment the dynamic content.  Next, we asked whether some correspondence might be made between the specific content of the events the participants experienced while viewing the episode, and the events they later recalled.  We labeled each recall event as matching the episode event with the most similar (i.e., most highly correlated) topic vector (Fig.~\ref{fig:model}G, Supp.\ Fig.~\matchmats).  This yielded a sequence of ``presented'' events from the original episode, and a (potentially differently ordered) sequence of ``recalled'' events for each participant.  Analogous to classic list-learning studies, we can then examine participants' recall sequences by asking which events they tended to recall first (probability of first recall~\citep{AtkiShif68, PostPhil65, WelcBurn24}; Fig.~\ref{fig:list-learning}A); how participants most often transitioned between recalls of the events as a function of the temporal distance between them (lag-conditional response probability~\citep{Kaha96}; Fig.~\ref{fig:list-learning}B); and which events they were likely to remember overall (serial position recall analyses~\citep{Murd62a}; Fig.~\ref{fig:list-learning}C). Some of the patterns we observed appeared to be similar to classic effects from the list-learning literature.  For example, participants had a higher probability of initiating recall with early events (Fig.~\ref{fig:list-learning}A) and a higher probability of transitioning to neighboring events with an asymmetric forward bias (Fig.~\ref{fig:list-learning}B). However, unlike what is typically observed in list-learning studies, we did not observe patterns comparable to the primacy or recency serial position effects (Fig.~\ref{fig:list-learning}C).  We hypothesized that participants might be leveraging meaningful narrative associations and references over long timescales throughout the episode.

@@

Clustering scores are often used by memory researchers to characterize how people organize their memories of words on a studied list~\citep{PolyEtal09}.  We defined analogous measures to characterize how participants organized their memories for episodic events (see \textit{Methods} for details).  Temporal clustering refers to the extent to which participants group their recall responses according to encoding position.  Overall, we found that sequentially viewed episode events tended to appear nearby in participants' recall event sequences (mean clustering score: 0.732, SEM: 0.033).  Participants with higher temporal clustering scores tended to exhibit better overall memory for the episode, according to both Chen et al. (2017)~\cite{ChenEtal17}'s hand-counted numbers of recalled scenes from the episode (Pearson's $r(15) = 0.49,~p = 0.046$) and the numbers of episode events that best-matched at least one recall event (i.e., model-estimated number of events recalled; Pearson's $r(15) = 0.59,~p = 0.013$).  Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity~\citep{MannKaha12}.  We found that participants tended to recall semantically similar episode events together (mean clustering score: 0.650, SEM: 0.032), and that semantic clustering score was also related to both hand-counted  (Pearson's $r(15) = 0.65,~p = 0.005$) and model-estimated (Pearson's $r(15) = 0.58,~p = 0.015$) numbers of recalled events.

\begin{figure}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_distinctiveness}
  \caption{\small \textbf{Novel content-based metrics of naturalistic memory: precision and distinctiveness.} \textbf{A.} The episode-recall correlation matrix for a representative participant (P17), chosen for their large number of recall events (for analogous figures for other participants, see Supp.\ Fig.~\corrmats).  The yellow boxes highlight the maximum correlation in each column.  The example participant's overall precision score was computed as the average across the (Fisher $z$-transformed) correlation values in the yellow boxes.  Their distinctiveness score was computed as the average (over recall events) of the $z$-scored (within column) event precisions. \textbf{B.} The across-participants (Pearson's) correlation between precision and hand-counted number of recalled scenes.  \textbf{C.} The correlation between distinctiveness and hand-counted number of recalled scenes. \textbf{D.} The correlation between precision and the number of recalled episode events, as determined by our model. \textbf{E.} The correlation between distinctiveness and the number of recalled episode events, as determined by our model.}
  \label{fig:precision-distinctiveness}
\end{figure}

@@

\begin{figure}[tp]
  \centering
  \vspace*{-1cm}
  \includegraphics[width=.7\textwidth]{figs/precision_distinctiveness_detail}
  \caption{\small \textbf{Precision reflects the completeness of recall, whereas distinctiveness reflects recall specificity.} \textbf{A.} Recall precision by episode event.  Grey violin plots display kernel density estimates for the distribution of recall precision scores for a single episode event.  Colored dots within each violin plot represent individual participants' recall precisions for the given event.  \textbf{B.} Recall distinctiveness by episode event, analogous to Panel A.  \textbf{C.} The set of ``Narrative Details'' episode annotations~\citep{ChenEtal17} comprising an example episode event (22) identified by the HMM.  Each action or feature is highlighted in a different color.  \textbf{D.} Sentences comprising the most precise (P17) and least precise (P6) participants' recalls of episode event 21.  Descriptions of specific actions or features reflecting those highlighted in Panel B are highlighted in the corresponding color.  The text highlighted in gray denotes a (rare) false recall.  The unhighlighted text denotes correctly recalled information about other episode events.  \textbf{E.} The sets of ``Narrative Details'' episode annotations~\citep{ChenEtal17} for scenes comprising episode events described by the example participants in Panel F.  Each event's text is highlighted in a different color.  \textbf{F.} The sentences comprising the most distinctive (P9) and least distinctive (P6) participants' recalls of episode event 21.  Sections of recall describing each each episode event in Panel E are highlighted with the corresponding color.}
  \label{fig:precision-detail}
\end{figure}

@@

The preceding analyses sought to characterize how participants' recountings of individual episode events captured the low-level details of each event.  Next, we sought to characterize how participants' recountings of the full episode captured its high-level essence---i.e., the shape of the episode's trajectory through word embedding (topic) space.  To visualize the essence of the episode and each participant's recall trajectory~\citep{HeusEtal18a}, we projected the topic proportions matrices for the episode and recalls onto a shared two-dimensional space using Uniform Manifold Approximation and Projection (UMAP)~\citep{McInEtal18}.  In this lower-dimensional space, each point represents a single episode or recall event, and the distances between the points reflect the distances between the events' associated topic vectors (Fig.~\ref{fig:trajectory}). In other words, events that are nearer to each other in this space are more semantically similar, and those that are farther apart are less so.

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/trajectory}
\caption{\small \textbf{Trajectories through topic space capture the dynamic content of the episode and recalls.}  All panels: the topic proportion matrices have been projected onto a shared two-dimensional space using UMAP.  \textbf{A.} The two-dimensional topic trajectory taken by the episode of \textit{Sherlock}.  Each dot indicates an event identified using the HMM (see \textit{Methods}); the dot colors denote the order of the events (early events are in red; later events are in blue), and the connecting lines indicate the transitions between successive events.  \textbf{B.} The average two-dimensional trajectory captured by participants' recall sequences, with the same format and coloring as the trajectory in Panel A.  To compute the event positions, we matched each recalled event with an event from the original episode (see \textit{Results}), and then we averaged the positions of all events with the same label.  The arrows reflect the average transition direction through topic space taken by any participants whose trajectories crossed that part of topic space; blue denotes reliable agreement across participants via a Rayleigh test ($p < 0.05$, corrected).  For additional detail see \textit{Methods} and Supplementary Figure~\arrows.  \textbf{C.} The recall topic trajectories (gray) taken by each individual participant (P1--P17).  The episode's trajectory is shown in black for reference.  Here, events (dots) are colored by their matched episode event (Panel A).}
\label{fig:trajectory}
\end{figure}

Visual inspection of the episode and recall topic trajectories reveals a striking pattern.  First, the topic trajectory of the episode (which reflects its dynamic content; Fig.~\ref{fig:trajectory}A) is captured nearly perfectly by the averaged topic trajectories of participants' recalls (Fig.~\ref{fig:trajectory}B).  To assess the consistency of these recall trajectories across participants, we asked: given that a participant's recall trajectory had entered a particular location in the reduced topic space, could the position of their next recalled event be predicted reliably?  For each location in the reduced topic space, we computed the set of line segments connecting successively recalled events (across all participants) that intersected that location (see \textit{Methods}, Supp.\ Fig.~\arrows).  We then computed (for each location) the distribution of angles formed by the lines defined by those line segments and a fixed reference line (the $x$-axis).  Rayleigh tests revealed the set of locations in topic space at which these across-participant distributions exhibited reliable peaks (blue arrows in Fig.~\ref{fig:trajectory}B reflect significant peaks at $p < 0.05$, corrected).  We observed that the locations traversed by nearly the entire episode trajectory exhibited such peaks.  In other words, participants' recalls exhibited similar trajectories to each other that also matched the trajectory of the original episode (Fig.~\ref{fig:trajectory}C).  This is especially notable when considering the fact that the number of HMM-identified recall events (dots in Fig.~\ref{fig:trajectory}C) varied considerably across people, and that every participant used different words to describe what they had remembered happening in the episode.  Differences in the numbers of recall events appear in participants' trajectories as differences in the sampling resolution along the trajectory.  We note that this framework also provides a means of disentangling classic ``proportion recalled'' measures (i.e., the proportion of episode events described in participants' recalls) from participants' abilities to recapitulate the episode's essence (i.e., the similarity between the shapes of the original episode trajectory and that defined by each participant's recounting of the episode).
